Radial and rotational velocities of young brown dwarfs and very low-mass stars in the Upper Scorpius OB association and the $\rho$~Ophiuchi cloud core

### 2006-06-20 BD_radvel/kurosawa/kurosawa.tex
typos tidied up - th
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@747 25464f21-6099-402c-b30a-640f1d40d38b

We present the results of a radial velocity (RV) survey of 14 brown
dwarfs (BDs) and very low-mass (VLM) stars in the Upper Scorpius OB
association (UScoOB) and 3 BD candidates in the $\rho$~Ophiuchi dark
cloud core. We obtained high-resolution echelle spectra at the VLT using
Ultraviolet and Visual Echelle Spectrograph (UVES) at two different
epochs for each object, and measured the shifts in their RVs to
identify candidates for binary/multiple systems in the sample. The
average time separation of the RV measurements is 21.6~d, and our
survey is sensitive to the binaries with separation $<0.1$~au.  We
found 4 out of 17 objects (or $24^{+16}_{-13}$~per~cent by fraction)
show a significant RV change in 4--33~d time scale, and are 
considered as binary/multiple `candidates.'  We found no double-lined
spectroscopic binaries in our sample, based on the shape of
cross-correlation curves. The RV dispersion of the objects in UScoOB
is found to be very similar to that of the BD and VLM stars in
Chamaeleon~I (Cha~). We also found the distribution of the mean
rotational velocities ($v\,\sin i$) of the 
UScoOB objects is similar to that of the Cha~I, but the dispersion of
$v\,\sin i$ is much larger than that of the Cha~I objects.

@@

The extensive imaging surveys provide excellent observational constraints
on wider BD+BD binaries, but it is now necessary to search for shorter
period BD+BD binaries systematically. Binaries with the separation
of less than 1~au are not resolved by current imaging techniques,
but will be detectable as spectroscopic binaries, providing the mass-ratio
is not too extreme, and velocity separation is large enough. The first
BD+BD spectroscopic binary, PPl~15  \citep{basri:1999}, showed a
double-peaked cross-correlation function with a maximum velocity separation
of $>70\,\kmps$. The binary was found to have an eccentric orbit
($e=0.4$) with a period of $\sim5.8\,\mathrm{d}$. \citet{basri:1999}
suggested that the formation process of substellar objects is biased
towards smaller separation binaries based on the short period of PPl~15
and the lack of Pleiades BD binaries with separations $>40\,\mathrm{au}$.
Note that the median separation of binaries with solar-type primaries
is 30~au \citep{duquennoy:1991}. Pioneering works on the RVs
of BDs and VLMs are presented by \citet{guenther:2003}, \cite{kenyon:2005}
and \citet{joergens:2006} who found a several binary candidates;
however, the orbital parameters and masses of binaries remains
unknown because the follow-up spectroscopic monitoring is lacking
or still being undertaken. In addition to the follow-up observations,
the number of BDs and VLM star binary candidates needs to be increased
in order to have better statistics on short-period binary parameters.

@@

Our immediate aim is to identify spectroscopic and close BD/VLM binaries
using the high-resolution echelle spectroscopy at two epochs. This
experiment is sensitive to VLM binaries with separations of $<0.1$~au
which corresponds to a period of $\sim10\,\mathrm{d}$. A larger sample
of candidates will enable us to measure the binary fraction of these
short-period/close binaries (once confirmed), and address whether
there is a significant population of 'hidden' VLM companions. The
long term goal of this project to follow up the binary candidates
found in this paper by spectroscopically monitoring them over different
time scales, enabling us to obtain the radial velocity curves and
their minimum masses.

In Section~\ref{sec:observation-reduction}. we describe the
observations and the data reduction. The results of radial velocity
and rotational velocity ($v\,\sin i$) measurements are presented in
Section~\ref{sec:Results}. We discuss the binary/multiplicity fraction
indicated by our RV survey in Section~\ref{sec:Binary-fraction}, and
give our conclusions in Section~\ref{sec:Conclusions}.

@@

We obtained high-resolution spectra with the Kueyen telescope of VLT
(Cerro Parnal, Chile) using the UVES echelle spectrograph. The
observations were carried out between 2004 April 5 and 2004 May 17 in
the service mode. For each object, spectra were obtained at two different
epochs separated by 4--33~d. For each object at a
given night, two separate spectra are obtained consecutively. This
allows us to derive more reliable uncertainty estimates in the RV
values of our targets (c.f.~\citealt{joergens:2006}).  The data were
obtained using the red arm of UVES spectrograph with two mosaic CCDs
(EEV + MIT/LL with 2k$\times$4k pixels). The wavelength coverage of
6708 -- 10,250~\AA\, and the spectral resolution $R\approx40,000$ were
used. The slit width and length of 1'' and 12'' were used respectively
with a typical seeing of 0.8''.

@@

\caption{Summary of known properties of the targets from literatures: a.~\citet{luhman:1999}
(original list for $\rho$~Oph), b.~\citet*{ardila:2000} (original
list for UScoOB), c.~\citet*{wilking:1999}, d.~\citet{muzerolle:2003},
e.~\citet{kraus:2005}, and f.~\citet*{mohanty:2005}. }

@@

The result of the heliocentric RV measurements (from two epochs for
each object) is summarised in Table~\ref{tab:summary_results} along
with the uncertainties. The table also lists the uncertainties in
the `relative' radial velocities ($\sigma_{\mathrm{RRV}}$) which
is the uncertainty of RV with respect to the template star. 
To determine the $\sigma_{\mathrm{RRV}}$, we have used the following
procedure: (1)~add random gaussian deviate noise to the object spectra using the
corresponding variance spectra, (2)~compute the cross-correlation
curve using the spectra with added noise and the template, (3)~measure
the RV by locating the peak, (4)~repeat 1--3 for 100 times, and compute the standard
deviation of the 100 RV values. This should provide a good estimate of
uncertainties in RVs cased by the uncertainties in the flux levels.
We have also used an alternative method of estimating
$\sigma_{\mathrm{RRV}}$ values as described in
(c.f.~\citealt{joergens:2006}). The standard deviations of the mean of
the two independent RV measurements obtained from the two consecutive
spectra taken in the same night were assigned as the uncertainties
($\sigma_{\mathrm{RRV}}$). Although not shown here, we found that the
results from the two methods agree each other quite well. 
Note that for finding a shift in the radial velocities from two
different epochs, only $\sigma_{\mathrm{RRV}}$ is important since one
only requires the relative (with respect to a template) RV values.
The `average' radial velocities $\left (\overline{\mathrm{RV}} \right )$ of the
two epochs are also given in the same table.

For each object and for each RV measurement, the deviations ($\Delta\mathrm{RV}$)
from the average RV are computed and summarized in Fig.~\ref{fig:radvel01}
along with their uncertainties ($\sigma_{\mathrm{RRV}}$) in order
to aid the identification of multiplicity candidates. Note that in
computing $\Delta\mathrm{RV}$ we do not require the knowledge of
absolute radial velocities, but only the relative
velocities (with respect to a template). 
Alternatively, we also computed the cross-correlation functions of
spectra of a same object (instead of using the spectra of the template
LHS~49) from two different epochs in order to find
$\Delta\mathrm{RV}$. The resulting cross-correlation functions were
found to be too noisy (due to the relatively low S/N in the object
spectra), and the peak positions of the cross-correlation curves were
not reliably located.

@@

Finally, the histogram of $\overline{\mathrm{RV}}$ for the objects in UScoOB
is given in Fig.~\ref{fig:vrad_histrogram}. The total number
of the objects is 14. Note that USco~121 is excluded from the graph
since it is identified as a non-member of the UScoOB association based
on the RV value (see Table~\ref{tab:summary_results}).  \citet{muzerolle:2003}
also found it to be a likely non-member based on the radial velocity
and the low lithium abundance. The distribution of the RVs in the
figure was fitted by a gaussian function. We found that the standard
deviation and the peak position of the radial velocity distribution
are $1.0\,\kmps$ and $-6.3\,\kmps$ respectively. The former is very
similar to the standard deviation (0.9~$\kmps$) of the radial velocity
distribution of 9 BDs and VLM objects in Cha~I found by \citet{joergens:2006b}.
They also studied the radial velocity distribution of more massive
25 T~Tauri stars in Cha~I, and found the standard deviations ($1.3\,\kmps$)
is not significantly different from that of the brown dwarfs and the
very low-mass objects. Unfortunately, we do not have the radial velocity
measurements of the higher mass counter parts (T~Tauri stars) in
Upper Sco OB association. This is planned for a near future investigation
as this is important for the study of the mass dependency of the kinematics
in a young stellar cluster.

@@

In Section~\ref{sub:Radial-velocities}
(see Fig.~\ref{fig:chisq-test}), we found the histogram of the
$\chi^{2}$ probability $p$, for the RV data fit assuming constant RV,
reasonably matches the expected distribution.  This leas us to
conclude that the uncertainty estimates in RV values are 
reasonable.  There are two possible
source of errors: (1)~the RV shifts caused by surface/atmospheric
activity, and (2)~the RV shifts due to the limit of the stability of
autoguiding of the telescope.  \citet{joergens:2006} found that (1) is
negligibly small for young VLM objects with $M_{*}<0.12\,M_{\sun}$.
The size of RV errors caused by (2) is expected to be $<1\,\kmps$
\citep{bailer-jones:2004}. Note that \citet{maxted:2005} found an
additional systematic error was not needed to be added to the RV
uncertainties of \citet{joergens:2006} who used the same instruments
as ours, and used the same method of estimating the
uncertainties. Hence, we expect the systematic error to be much
smaller than $1\,\kmps$ if one is required.

An alternative explanation for the RV variations found in the binary
candidates above is a type of stellar pulsations found by
\citet{palla:2005}. They studied the non-adiabatic, linear
instability of very VLM stars and BD during the deturium burning phase
in the core, and found unstable fundamental modes in the time-scale
between $\sim 1$ and $\sim 5$~h for the object mass from $0.02$ and
$0.1\,\Msun$.  By using their pulsation periods and by assuming that the
pulsation amplitudes is 5~per~cent of the radius of the objects, we
find that the expected RV variation can be $>6\,\kmps$.
As mentioned in Section~\ref{sec:observation-reduction}, for each object at a
given night, two separate spectra are obtained consecutively (with
time separations of $\sim 0.5$~h.). Before co-adding the two spectra
together, we have measured the RV from each spectrum to check if there
is a large jump in the RV measurements in short
time-scale. The size of the shifts in the RV values measured in the two
consecutive spectra are found in between $0.02\pm0.10$ (USco~128)
and  $2.04\pm1.67\,\kmps$ (USco~66), and the average size of the
shifts $0.36\,\kmps$, which is much smaller than the
expected RV variation based on the models of \citet{palla:2005};
hence, we found little indication of the pulsations in our sample.

@@

We found 4 (USco~112, USco~128, USco~40 and USco~55) out of 17 objects as our 
binary/multiple candidates. This corresponds to the binary fraction of
$24^{+16}_{-13}$~per~cent for the binary separation $a<0.1$~au.  
The recent high-resolution imaging survey of brown dwarfs and very
low-mass objects (M5.5--M7.5 ) in the UScoOB by \citet{kraus:2005} (which was
not known to authors at the time of our observation: April--May, 2004)
confirms that USco~55 and USco~66 are multiple systems, and USco~109
is most likely a multiple system.  Their projected separations are
$a>4.0$~au which is well above our sensitivity limit of
0.1~au; therefore, it is not surprising that we did not identify USco~66 and
USco~109 as multiples. Interestingly, we also found USco~55 as 
a candidate for a multiple system. This may indicate that it is  
a triple system with one of objects located within 0.1~au from the
primary.  In addition, we identify USco~112 and 128 as binary
candidates, but they did not find them as multiples. Further they found USco~67, 75,
130 and 132 as non-multiple, and so did we.

>>>>

We present the results of a radial velocity (RV) survey of 14 brown
dwarfs (BDs) and very low-mass (VLM) stars in the Upper Scorpius OB
association (UScoOB) and 3 BD candidates in the $\rho$~Ophiuchi dark
cloud core. We obtained high-resolution echelle spectra at the
Very Large Telescope using Ultraviolet and Visual Echelle
Spectrograph (UVES) at two different epochs for each object,
and measured the shifts in their RVs to
identify candidates for binary/multiple systems in the sample. The
average time separation of the RV measurements is 21.6~d, and our
survey is sensitive to the binaries with separation $<0.1$~au.  We
found that  4 out of 17 objects (or $24^{+16}_{-13}$~per~cent by fraction)
show a significant RV change in 4--33~d time scale, and are 
considered as binary/multiple `candidates.'  We found no double-lined
spectroscopic binaries in our sample, based on the shape of
cross-correlation curves. The RV dispersion of the objects in UScoOB
is found to be very similar to that of the BD and VLM stars in
Chamaeleon~I (Cha~I). We also found the distribution of the mean
rotational velocities ($v\,\sin i$) of the 
UScoOB objects is similar to that of the Cha~I, but the dispersion of
$v\,\sin i$ is much larger than that of the Cha~I objects.

@@

The extensive imaging surveys provide excellent observational constraints
on wider BD+BD binaries, but it is now necessary to search for shorter
period BD+BD binaries systematically. Binaries with the separation
of less than 1~au are not resolved by current imaging techniques,
but will be detectable as spectroscopic binaries, providing the mass-ratio
is not too extreme, and velocity separation is large enough. The first
BD+BD spectroscopic binary, PPl~15  \citep{basri:1999}, showed a
double-peaked cross-correlation function with a maximum velocity separation
of $>70\,\kmps$. The binary was found to have an eccentric orbit
($e=0.4$) with a period of $\sim5.8\,\mathrm{d}$. \citet{basri:1999}
suggested that the formation process of substellar objects is biased
towards smaller separation binaries based on the short period of PPl~15
and the lack of Pleiades BD binaries with separations $>40\,\mathrm{au}$.
Note that the median separation of binaries with solar-type primaries
is 30~au \citep{duquennoy:1991}. Pioneering work on the RVs
of BDs and VLMs are presented by \citet{guenther:2003}, \cite{kenyon:2005}
and \citet{joergens:2006} who found a several binary candidates;
however, the orbital parameters and masses of binaries remains
unknown because the follow-up spectroscopic monitoring is lacking
or is still being undertaken. In addition to the follow-up observations,
the number of BDs and VLM star binary candidates needs to be increased
in order to have better statistics on short-period binary parameters.

@@

Our immediate aim is to identify spectroscopic and close BD/VLM binaries
using the high-resolution echelle spectroscopy at two epochs. This
experiment is sensitive to VLM binaries with separations of $<0.1$~au
which corresponds to a period of $\sim10\,\mathrm{d}$. A larger sample
of candidates will enable us to measure the binary fraction of these
short-period/close binaries (once confirmed), and address whether
there is a significant population of `hidden' VLM companions. The
long term goal of this project to follow up the binary candidates
found in this paper by spectroscopically monitoring them over different
time scales, enabling us to obtain the radial velocity curves and
their minimum masses.

In Section~\ref{sec:observation-reduction} we describe the
observations and the data reduction. The results of radial velocity
and rotational velocity ($v\,\sin i$) measurements are presented in
Section~\ref{sec:Results}. We discuss the binary/multiplicity fraction
indicated by our RV survey in Section~\ref{sec:Binary-fraction}, and
give our conclusions in Section~\ref{sec:Conclusions}.

@@

We obtained high-resolution spectra with the Kueyen telescope of the Very Large Telescope
(VLT--Cerro Parnal, Chile) using the UVES echelle spectrograph. The
observations were carried out between 2004 April 5 and 2004 May 17 in
the service mode. For each object, spectra were obtained at two different
epochs separated by 4--33~d. For each object on a
given night, two separate spectra were obtained consecutively. This
allows us to derive more reliable uncertainty estimates in the RV
values of our targets (c.f.~\citealt{joergens:2006}).  The data were
obtained using the red arm of UVES spectrograph with two mosaic CCDs
(EEV + MIT/LL with 2k$\times$4k pixels). The wavelength coverage of
6708 -- 10,250~\AA\, and the spectral resolution $R\approx40,000$ were
used. The slit width and length of 1'' and 12'' were used respectively
with a typical seeing of 0.8''.

@@

\caption{Summary of known properties of the targets from literature: a.~\citet{luhman:1999}
(original list for $\rho$~Oph), b.~\citet*{ardila:2000} (original
list for UScoOB), c.~\citet*{wilking:1999}, d.~\citet{muzerolle:2003},
e.~\citet{kraus:2005}, and f.~\citet*{mohanty:2005}. }

@@

The result of the heliocentric RV measurements (from two epochs for
each object) is summarised in Table~\ref{tab:summary_results} along
with the uncertainties. The table also lists the uncertainties in
the `relative' radial velocities ($\sigma_{\mathrm{RRV}}$) which
is the uncertainty of RV with respect to the template star. 
To determine the $\sigma_{\mathrm{RRV}}$, we have used the following
procedure: (1)~add random gaussian deviate noise to the object spectra using the
corresponding variance spectra, (2)~compute the cross-correlation
curve using the spectra with added noise and the template, (3)~measure
the RV by locating the peak, (4)~repeat 1--3 for 100 times, and compute the standard
deviation of the 100 RV values. This should provide a good estimate of
uncertainties in RVs cased by the uncertainties in the flux levels.
We also estimated the RV uncertainties using the standard deviations on the
mean of the two independent RV measurements obtained from the two
consecutive spectra taken on the same night (c.f.~\citealt{joergens:2006}).
We found that these estimates agreed well with those obtained from
the Monte Carlo method. The `average' radial velocities $\left (\overline{\mathrm{RV}} \right )$ of the
two epochs are also given in Table~\ref{tab:summary_results}.

For each object and for each RV measurement, the deviations ($\Delta\mathrm{RV}$)
from the average RV are computed and summarized in Fig.~\ref{fig:radvel01}
along with their uncertainties ($\sigma_{\mathrm{RRV}}$) in order
to aid the identification of multiplicity candidates. Note that in
computing $\Delta\mathrm{RV}$ we do not require the knowledge of
absolute radial velocities, but only the relative
velocities (with respect to a template). 
We also  cross-correlated object spectra from the two epochs 
(instead of using the spectra of the template LHS~49)  in order to find
$\Delta\mathrm{RV}$. The resulting cross-correlation functions were
found to be too noisy (due to the relatively low S/N in the object
spectra) for the  peak positions of the cross-correlation curves to be
reliably located.

@@

Finally, the histogram of $\overline{\mathrm{RV}}$ for the objects in UScoOB
is given in Fig.~\ref{fig:vrad_histrogram}. The total number
of the objects is 14. Note that USco~121 is excluded from the graph
since it is identified as a non-member of the UScoOB association based
on the RV value (see Table~\ref{tab:summary_results}).  \citet{muzerolle:2003}
also found it to be a likely non-member based on the radial velocity
and the low lithium abundance. The distribution of the RVs in the
figure was fitted by a gaussian function. We found that the standard
deviation and the peak position of the radial velocity distribution
are $1.0\,\kmps$ and $-6.3\,\kmps$ respectively. The former is very
similar to the standard deviation (0.9~$\kmps$) of the radial velocity
distribution of 9 BDs and VLM objects in Cha~I found by \citet{joergens:2006b}.
They also studied the radial velocity distribution of more massive
25 T~Tauri stars in Cha~I, and found the standard deviations ($1.3\,\kmps$)
is not significantly different from that of the brown dwarfs and the
very low-mass objects. Unfortunately, we do not have the radial velocity
measurements of the higher mass counter parts (T~Tauri stars) in
Upper Sco OB association. This is planned for  future investigation
since this is important for the study of the mass dependency of the kinematics
in a young stellar cluster.

@@

In Section~\ref{sub:Radial-velocities}
(see Fig.~\ref{fig:chisq-test}), we found the histogram of the
$\chi^{2}$ probability $p$, for the RV data fit assuming constant RV,
reasonably matches the expected distribution.  This leads us to
conclude that the uncertainty estimates in RV values are 
reasonable.  There are two possible
source of errors: (1)~the RV shifts caused by surface/atmospheric
activity, and (2)~the RV shifts due to the limit of the stability of
autoguiding of the telescope.  \citet{joergens:2006} found that (1) is
negligibly small for young VLM objects with $M_{*}<0.12\,M_{\sun}$.
The size of RV errors caused by (2) is expected to be $<1\,\kmps$
\citep{bailer-jones:2004}. Note that \citet{maxted:2005} found an
additional systematic error was not needed to be added to the RV
uncertainties of \citet{joergens:2006} who used the same instruments
as ours, and used the same method of estimating the
uncertainties. Hence, we expect the systematic error to be much
smaller than $1\,\kmps$ if one is required.

An alternative explanation for the RV variations found in the binary
candidates above is a type of stellar pulsations found by
\citet{palla:2005}. They studied the non-adiabatic, linear
instability of very VLM stars and BD during the deturium burning phase
in the core, and found unstable fundamental modes in the time-scale
between $\sim 1$ and $\sim 5$~h for the object mass from $0.02$ and
$0.1\,\Msun$.  By using their pulsation periods and by assuming that the
pulsation amplitudes is 5~per~cent of the radius of the objects, we
find that the expected RV variation can be $>6\,\kmps$.
As mentioned in Section~\ref{sec:observation-reduction}, for each object at a
given night, two separate spectra are obtained consecutively (with
time separations of $\sim 0.5$~h.). We have measured the RV from each spectrum to check if there
is a large jump in the RV measurements on a  short
time-scale. The size of the shifts in the RV values measured in the two
consecutive spectra are found in between $0.02\pm0.10$ (USco~128)
and  $2.04\pm1.67\,\kmps$ (USco~66), and the average size of the
shifts $0.36\,\kmps$, which is much smaller than the
expected RV variation based on the models of \citet{palla:2005};
hence, we found little indication of the pulsations in our sample.

@@

We found 4 (USco~112, USco~128, USco~40 and USco~55) out of 17 objects as our 
binary/multiple candidates. This corresponds to the binary fraction of
$24^{+16}_{-13}$~per~cent for the binary separation $a<0.1$~au.  
The recent high-resolution imaging survey of brown dwarfs and very
low-mass objects (M5.5--M7.5 ) in the UScoOB by \citet{kraus:2005} (which was
not known to authors at the time of our observation: April--May, 2004)
confirms that USco~55 and USco~66 are multiple systems, and USco~109
is most likely a multiple system.  Their projected separations are
$a>4.0$~au which is well above our sensitivity limit of
0.1~au; therefore, it is not surprising that we did not identify USco~66 and
USco~109 as multiples. Interestingly, we also found USco~55 as 
a candidate for a multiple system. This may indicate that it is  
a triple system with one of the objects located within 0.1~au of the
primary.  In addition, we identify USco~112 and 128 as binary
candidates, but they did not find them as multiples. Further they found USco~67, 75,
130 and 132 as non-multiple, as did we.
